---
title: "Predicting Survival From Colic"
author: "Kate O'Rourke"
date: "2024-09-04"
output: pdf_document
---

# Load Libraries

```{r}
library(tidyverse)
library(tidyr)
library(caret)
library(dplyr)
library(mice)
```

# Load Data

```{r}
colic_training <- read.delim("Data/horse-colic.data", sep=" ", header=FALSE)
colic_training$V29 <- NULL
colic_test <- read.delim("Data/horse-colic.test", sep=" ", header=FALSE)
```

## Merge the two datasets
```{r}
## Merge test and training datasets
colic_data <- rbind(colic_test, colic_training)
```

## Rename columns

```{r}
colic_data <- colic_data %>% 
  rename(
    surgery = V1,
    age = V2,
    hospital_id = V3,
    rectal_temperature = V4,
    pulse = V5,
    respiratory_rate = V6,
    extremity_temperature = V7,
    peripheral_pulse = V8,
    mucous_membrane = V9,
    capillary_refill_time = V10,
    pain_rating = V11,
    peristalsis = V12,
    abdominal_distension = V13,
    nasogastric_tube = V14,
    nasogastric_reflux = V15,
    nasogastric_reflux_pH = V16,
    rectal_exam_feces = V17,
    abdomen = V18,
    packed_cell_Volume = V19,
    total_protein = V20,
    abdominocentesis_appearance = V21,
    abdominocentesis_total_protein = V22,
    outcome = V23,
    surgical_lesion = V24,
    type_of_lesion1 = V25,
    type_of_lesion2 = V26,
    type_of_lesion3 = V27,
    cp_data = V28
    )
```

## Deal with ? or missing data
```{r}
# Replace "?" with NA
colic_data <-  replace(colic_data, colic_data=="?", NA)
colSums(is.na(colic_data))

# Remove columns where there are fewer than 70% of data points populated
colic_data <- colic_data[, -c(14:18, 21:22, 24:28) ]

# Remove hospital_id column
colic_data <- colic_data[, -3 ]

# Change levels in outcome to align with 1 being survived and 0 being did not survive
colic_data$outcome[colic_data$outcome == "1"] <- 1
colic_data$outcome[colic_data$outcome == "2"] <- 0
colic_data$outcome[colic_data$outcome == "3"] <- 0
```

# Determine training-test split
```{r}
## Written by Katherine S. Geist, PhD
## Merrimack College, Massachusetts
## Please do not distribute without attribution
## https://github.com/ksgeist

calcSplitRatio <- function(p = NA, df) {
  ## @p  = the number of parameters. by default, if none are provided, the number of columns (predictors) in the dataset are used
  ## @df = the dataframe that will be used for the analysis
  
  ## If the number of parameters isn't supplied, set it to the number of features minus 1 for the target
  if(is.na(p)) {
    p <- ncol(df) -1   ## COMMENT HERE
  }
  
  ## Calculate the ideal number of testing set
  test_N <- (1/sqrt(p))*nrow(df)
  ## Turn that into a testing proportion
  test_prop <- round((1/sqrt(p))*nrow(df)/nrow(df), 2)
  ## And find the training proportion
  train_prop <- 1-test_prop
  
  ## Tell us the results!
  print(paste0("The ideal split ratio is ", train_prop, ":", test_prop, " training:testing"))
  
  ## Return the size of the training set
  return(train_prop)
}

calcSplitRatio(p=NA, colic_data)
```

## Remove rows with NA values for surgery or outcome
```{r}
# Remove rows where surgery or outcome have NA value
colic_data <- colic_data %>%
  filter(
    !is.na(surgery),
    !is.na(outcome)
    )
```

```{r}
# Convert columns to numeric if containing numeric or ordered data with more than 2 values, otherwise convert to factor
colic_data[, c(3:5, 11:14)] <- sapply(colic_data[, c(3:5, 11:14)], as.numeric)
colic_data[, c(1:2, 5:11, 15)] <- lapply(colic_data[, c(1:2, 6:11, 15)], factor)
```

# Separate dataset into training and test sets
```{r}
training <- createDataPartition(colic_data$outcome,
                                         p = 0.73,
                                         list = FALSE,
                                         times = 1)

colic_training <- colic_data[training, ]
colic_test <- colic_data[-training, ]
```

# Impute values in datasets using MICE
```{r}
# MICE will be used to impute missing values in training and test sets
md.pattern(colic_training) # Get an idea of the pattern of missing data
imp <- mice(colic_training, seed = 223)
print(imp)
colic_training_imp <- complete(imp)

md.pattern(colic_test) # Get an idea of the pattern of missing data
imp <- mice(colic_test, seed = 223)
print(imp)
colic_test_imp <- complete(imp)
```

# One-hot encode factor variables
```{r}
# Training Set
onehot_encoder <- dummyVars(~ surgery + age + extremity_temperature + peripheral_pulse + mucous_membrane + capillary_refill_time + pain_rating + peristalsis,
                            colic_training_imp[, c("surgery", "age", "extremity_temperature", "peripheral_pulse", "mucous_membrane", "capillary_refill_time", "pain_rating", "peristalsis")],
                            levelsOnly = TRUE,
                            fullRank = TRUE)

onehot_enc_training <- predict(onehot_encoder,
                               colic_training_imp[, c("surgery", "age", "extremity_temperature", "peripheral_pulse", "mucous_membrane", "capillary_refill_time", "pain_rating", "peristalsis")])

colic_training_imp <- cbind(colic_training_imp, onehot_enc_training)

# Test Set
onehot_encoder <- dummyVars(~ surgery + age + extremity_temperature + peripheral_pulse + mucous_membrane + capillary_refill_time + pain_rating + peristalsis,
                            colic_test_imp[, c("surgery", "age", "extremity_temperature", "peripheral_pulse", "mucous_membrane", "capillary_refill_time", "pain_rating", "peristalsis")],
                            levelsOnly = TRUE,
                            fullRank = TRUE)

onehot_enc_training <- predict(onehot_encoder,
                               colic_test_imp[, c("surgery", "age", "extremity_temperature", "peripheral_pulse", "mucous_membrane", "capillary_refill_time", "pain_rating", "peristalsis")])

colic_test_imp <- cbind(colic_test_imp, onehot_enc_training)
```

# Normalize numeric columns
```{r}
# Function to normalize specific columns based on dataset and columns specified
normalize <- function(dataset, columns) {
  scale(dataset[, columns],
  center = apply(dataset[, columns], 2, mean),
  scale = apply(dataset[, columns], 2, sd)
  )
}
columns <- c(3:5, 12:14)
colic_training_imp[, columns] <- normalize(colic_training_imp, columns)
colic_test_imp[, columns] <- normalize(colic_test_imp, columns)
```

# Train a dense neural network to predict outcome - NOT YET EDITED
```{r}
# Unlist should include all numeric
training_features <- array(data = unlist(training_set[, -columns]),
                           dim = c(nrow(training_set), 35))
training_labels <- array(data = unlist(training_set[, 15]),
                         dim = c(nrow(training_set)))
test_features <- array(data = unlist(test_set[, -columns]),
                       dim = c(nrow(test_set), 35))
test_labels <- array(data = unlist(test_set[, 15]),
                     dim = c(nrow(test_set)))

use_virtualenv("my_tf_workspace")

model <- keras_model_sequential(list(
  layer_dense(units = 50, activation = "relu"),
  layer_dense(units = 25, activation = "relu"),
  layer_dense(units = 1, activation = "sigmoid")
))

compile(model,
        optimizer = "adam",
        loss = "binary_crossentropy",
        metrics = "accuracy")

history <- fit(model, training_features, training_labels,
               epochs = 40, batch_size = 512, validation_split = 0.33)
```

# Evaluating model

```{r}
predictions <- predict(model, test_features)
test_set$p_prob <- predictions[, 1]
head(predictions, 10)
```

# ROC Curve

```{r}
over_threshold <- test_set[test_set$p_prob >= 0.5, ]
fpr <- sum(over_threshold$booking_status==0)/sum(test_set$booking_status==0)
fpr
tpr <- sum(over_threshold$booking_status==1)/sum(test_set$booking_status==1)
tpr

roc_data <- data.frame(threshold=seq(1,0,-0.01), fpr=0, tpr=0)
for (i in roc_data$threshold) {
  over_threshold <- test_set[test_set$p_prob >= i, ]
  fpr <- sum(over_threshold$booking_status==0)/sum(test_set$booking_status==0)
  roc_data[roc_data$threshold==i, "fpr"] <- fpr
  tpr <- sum(over_threshold$booking_status==1)/sum(test_set$booking_status==1)
  roc_data[roc_data$threshold==i, "tpr"] <- tpr
}
ggplot() +
  geom_line(data = roc_data, aes(x = fpr, y = tpr, color = threshold), size = 2) +
  scale_color_gradientn(colors = rainbow(3)) +
  geom_abline(intercept = 0, slope = 1, lty = 2) +
  geom_point(data = roc_data[seq(1, 101, 10), ], aes(x = fpr, y = tpr)) +
  geom_text(data = roc_data[seq(1, 101, 10), ],
            aes(x = fpr, y = tpr, label = threshold, hjust = 1.2, vjust = -0.2))
```

# Accuracy

```{r}
test_results <-
  test_set %>%
  select(booking_status) %>%
  bind_cols(
    data.frame(p_1 = predictions)
  )

threshold <- 0.5

test_results$predictions <- ifelse(test_results$p_1 >= threshold, 1, 0)
tp <- nrow(test_results[test_results$booking_status==1 & test_results$predictions==1, ])
fp <- nrow(test_results[test_results$booking_status==0 & test_results$predictions==1, ])
tn <- nrow(test_results[test_results$booking_status==0 & test_results$predictions==0, ])
fn <- nrow(test_results[test_results$booking_status==1 & test_results$predictions==0, ])

accuracy <- (tp + tn) / (tp + fp + tn + fn)
accuracy
```

# AUC

```{r}
auc <- auc(x = roc_data$fpr, y = roc_data$tpr, type = "spline")
auc
```

# Calibration Curves

```{r}
in_interval <- test_set[test_set$p_prob >= 0.45 & test_set$p_prob <= 0.55, ]
```

```{r}
nrow(in_interval[in_interval$booking_status==1, ])/nrow(in_interval)
```

```{r}
calibration_data <- data.frame(bin_midpoint=seq(0.05,0.95,0.1),
observed_event_percentage=0)
for (i in seq(0.05,0.95,0.1)) {
  in_interval <- test_set[test_set$p_prob >= (i-0.05) & test_set$p_prob <= (i+0.05), ]
  oep <- nrow(in_interval[in_interval$booking_status==1, ])/nrow(in_interval)
  calibration_data[calibration_data$bin_midpoint==i, "observed_event_percentage"] <- oep
}
ggplot(data = calibration_data, aes(x = bin_midpoint, y = observed_event_percentage)) +
  geom_line(size = 1) +
  geom_abline(intercept = 0, slope = 1, lty = 2) +
  geom_point(size = 2) +
  geom_text(aes(label = bin_midpoint), hjust = 0.75, vjust = -0.5)
```